Markov Link Method for calibrating without joint measurement, including the case of destructive measurements

### 2018-04-20 trygve/paper.tex
New examples for MLM

\title{Markov Link Method for combining knowledge from destructive measurements}
\author{Jackson Loper}

@@

\begin{abstract}
When many different measurement tools are applied, how can we best combine information from the different different tools?  If we can apply multiple measurement tools to the same specimen, we can at least begin to understand how the tools they are related.  We call this ``joint measurement.''  What can we do if joint measurement is unavailable?  In this paper we propose the Markov Link Method: a simple assumption that can make it possible to obtain joint distributions linking tools together, even when joint measurement isn't possible.
\end{abstract}

The modern setting is rife with experimental methodologies, and it can be very frustrating to understand how the output of these methods relate to one another.  This is particularly difficult if joint measurement is unavailable, i.e.\ it is impossible or impractical to observe a single specimen with multiple methodologies.  For example, after processing a neuron with a given single-cell rna sequencing technique, we cannot generally go back and process that same neuron with a different technique.   This makes it seemingly impossible to calibrate the techniques against each other.  By contrast, it is easy to calibrate thermometers against each other, because we can simply measure the same water with two different thermometers and see how the measurements relate.  Without joint measurement, this isn't possible.

The absence joint measurement gives rise to a number of problems:

\begin{itemize}

\item Different tools may teach us different things, and we need to combine tools
to learn everything we can.  Let $A,B$ denote two different quantities of interest about cells (e.g. morphology and gene expression).   Let us say one tool enables us to measure $A$ and another tool to measure $B$.  We would like
to be able to say something about what morphologies are associated with what gene expressions.

\item Two labs may used different methods, but may wish to pool results.  Each lab looks at the other
lab's data and asks themselves "what would their data have looked like if they had used our method?"  When
they find they don't know the answer, it greatly complicates collaboration.

\item One may attempt to cluster the data.  It is simple to cluster the data from one technique, and cluster the data
from another technique -- but how can we know how those clusters correspond to one another?

\end{itemize}

In this work we propose the Markov Link Method (MLM) as a way to get useful and statistically rigorous bounds on the relationship between two measurement tools even when joint measurements are unavailable, hopefully yielding insight for the kinds of difficulties illustrated above.   In particular, we consider the case that we can obtain samples from a variety of subpopulations.  If we can assume that the \emph{relationship} between the measurements is the same in each subpopulation, we can use these samples to learn something about that relationship -- even though we never see both measurements for any single member of the population.

@@

This work stands on the shoulders of a long history of relating probabilistic assumptions to probabilistic inequalities.  Much of this literature comes from research into causality.  For example, in \cite{bonet2001instrumentality} Bonet uses polytopes not unlike the ones seen here to explore whether a variable can be used as an instrument.  The famous Clauser-Horne-Shimony-Holt inequality was designed to help answer causality questions in quantum physics, but it also sheds light on what distributions are consistent with certain assumptions \cite{clauser1969proposed}.  Indeed the physics literature has contributed many key inequalities (cf. \cite{chaves2014inferring}, \cite{kela2017semidefinite}, and the references therein).  Perhaps the closest work to this one would be \cite{makarov1982estimates}, which used two marginal distributions to get bounds on a property of the joint distribution (namely the distribution of the sum).  We build on this approach, both by using subpopulations to considerably refine our estimates and by considering the entire space of possible joint distributions instead of simply a particular aspect of the joint.

@@

Our motivation for this problem arose from looking at Allen Institute cell-type assignment of cells, performed using two different experimental techniques (also called experimental ``modalities'').  The only thing connecting the two modalities was that each modality had a variety of samples drawn from a variety of sub-populations, and the same sub-populations were used for both modalities.  Thus our input was two tables: (technique I cell-type $\times$ sub-population), and (technique II cell-type $\times$ sub-population) (Figure \ref{fig:alleninput}).  A key question immediately appears from these tables: does the way in which we have clustered cells into different types under technique I line up nicely with the way in which we have clustered cells under technique II?  The Markov Link Method gives us a way of addressing this question that is completely agnostic about the underlying genetic markers that were used to produce the clusters.  Our output was a set $\hat \Theta$ of possible ways that technique I cell-types could line up against the technique II cell-types.  That is, each element of $\hat \Theta$ was itself a table of the form (technique I cell-type $\times$ technique II cell-type).

>>>>

\title{Markov Link Method for calibrating without joint measurement, including the case of destructive measurements}
\author{Jackson Loper}

@@

\begin{abstract}
A proliferation of new experimental tools has left a serious gap: calibration. Two thermometers can be calibrated against each other by simply measuring the same bodies of water with both thermometers, but the problem is much harder for many modern tools.  One common problem is that we do not have measurements from the same ``body of water'' for both tools.  We propose the Markov Link Method (MLM) as a way to overcome this difficulty.   This method produces consistent estimators that tightly bound the calibration, i.e. the conditional distribution of one tool's measurement given another tool's measurement. It achieves this without any measurement data from both tools applied to the same ``bodies of water.'' Moreover, MLM makes zero assumptions about what calibrations we might expect to see, instead applying a subpopulation-based conditional independence assumption.   We evaluate MLM on a pair of single-cell RNA techniques, obtaining precise calibrations between the tools as well as more accurate models for each tool separately.
\end{abstract}

The modern setting is rife with experimental measurement tools, and it can be very frustrating to understand how the output of these tools relate to one another.  This problem is known as ``calibration'' or ``zeroing'' \cite{bipm2008international}.  A calibration tells us what readings we should expect from one tool, given the reading we obtained from another tool.   Calibration additionally must give uncertainty bounds for how much we can trust those expectations.  Calibration between measurement tools allows us to combine experimental results from different labs and different methodologies into larger scientific theories.

Formally, a calibration is simply a conditional distribution.  As input, this conditional distribution takes the measurement results from one tool on a particular specimen.  As output, it yields a distribution on what a second tool would measure on the same specimen.  One way to learn the calibration is to measure the same specimens under both tools.
We call this ``joint measurement.''  Unfortunately, calibrations are often required even when joint measurement is unavailable.  For example, if the measurement tool alters the specimen being measured, joint measurement is simply impossible.  In other cases, it may be expensive or impractical.

We here propose the Markov Link Method (MLM) to estimate calibrations between tools.  The MLM can be trained without any joint measurement.  The key idea is to use multiple subpopulations of specimens.  If each subpopulation captures a different slice of the overall population, we can obtain tight bounds on the true calibration.  This is true even if each subpopulation is highly heterogeneous.  By integrating information from all the subpopulations we can make rigorous deductions about what the calibration might be.  MLM also gives suggestions about which further subpopulations might be helpful to study in order to further refine our knowledge of the true calibration.  The method can also be naturally extended to calibration distributions among many tools.

Some example applications include:

\begin{itemize}
    \item Quality control for manufacturing.   The surest way to test the reliability of a part is to construct a machine that pushes the part until it breaks.  However, how can we test the reliability of the machine that performs the test?  In each test run there will be some variability induced by the machine itself, which induces a measurement error  In practice, some kind of assumptions about part homogeneity are used to approximate this error (cf.\ \cite{de2005gauge}).  However, if we have multiple testing machines we can use the MLM to obtain a calibration between the machines, even though we can never test the same part with both machines.  This enables us to bound the overall measurement error.

    \item Combining knowledge across experimental modalities: morphology and transcriptomics.  There are different ways to think about the different types of cells in an organism.  A traditional approach is to classify cells based on what they look like (cf.\ \cite{steinman1973identification,bloomfield1982physiological}).  A more modern approach is to assay the cell's transcriptome (cf.\ \citep{tasic2017shared}).  Unfortunately, modern high-resolution cell photography and single-cell sequencing technologies are both destructive.  As a result, we can't get both kinds of data for the same specimens.  For cells native to regions full of diverse cell-types, finding a correspondence is a real problem.  The result is two completely independent classifications of cells, one for each way of looking at the cell.  MLM allows us to estimate the relationship between those two classification systems, yielding a wholistic understanding of the different types of cells.

    \item Cancer treatment efficacy prediction.  Starting from in-vivo human cancers, many cell-lines have been cultured over the years.  These cell cultures live indefinitely on plates.  Many experiments have been performed to see how these cancer cells respond to treatment.  However, if a treatment work on a particular cultured cell-line, what can we say about which kinds of cancer might respond well to that treatment?  Coarse side-information such as original cancer location is often available for both in-vivo and cultured cells, but this is often a surprisingly weak signal.  Cell transcriptomes provides much more specific information about the cancer, and thus, in theory, what treatments might be appropriate (cf.\ \cite{cieslik2018cancer}).  However, we know that cultured cell-lines look quite different from in-vivo cells (cf. \cite{imamura2015comparison,haibe2013inconsistency}).  Moreover, we have very little joint measurement between the human cells which originally gave rise to a cell culture and the cells that survived to beceome the cell line.  The Markov Link method can leverage the common side-information together with separate transcriptome information to produce a fine-grained correspondence between in-vivo and cultured cells.  This correspondence can used to propose new treatments for cancers.

    \item Text/image correspondence.  Automatic image captioning is an ongoing effort in machine learning (cf.\ \cite{srivastava2018survey}).  There are three types of data available to help develop such algorithms: text-only data, image-only data, and paired-text-and-image data.  Obviously the last kind is the most useful for automatic image captioning, but there is much less of it.  The Markov Link Method suggests one way to use the more plentiful text-only and image-only data.  We can first apply classic machine learning techniques to get coarse labels for both kinds of data.  Using this side-information to identify subpopulations, the MLM can then deduce a fine-grained correspondence between text and images by combining information from across all the subpopulations.

    \item Replication crisis and lab effects.  Replicating a published study is not always an easy thing to do.  This difficulty is commonly attributed to selective publication bias, bad design, poor description of methods, and even outright fraud \cite{baker2016reproducibility}.  A calibration would allow us to understand this problem in detail.  If two labs perform identical experiments and get different data, that does not mean we need to throw out both datasets.  Instead, we can use MLM to calibrate the tools.  Once the tools are properly calibrated, we can combine both datasets.  Unlike other tools to deal with lab or batch effects (e.g. \cite{crow2018characterizing,johnson2007adjusting}), MLM makes zero assumptions about what calibrations we might expect.
\end{itemize}

@@

This work stands on the shoulders of a long history of relating probabilistic assumptions to probabilistic inequalities.  Much of this literature comes from research into causality.  For example, in \cite{bonet2001instrumentality} Bonet uses polytopes not unlike the ones seen here to explore whether a variable can be used as an instrument.  The famous Clauser-Horne-Shimony-Holt inequality was designed to help answer causality questions in quantum physics, but it also sheds light on what distributions are consistent with certain assumptions \cite{clauser1969proposed}.  Indeed the physics literature has contributed many key inequalities (cf. \cite{chaves2014inferring}, \cite{kela2017semidefinite}, and the references therein).  Perhaps the closest work to this one would be \cite{makarov1982estimates}, which used two marginal distributions to get bounds on a property of the joint distribution (namely the distribution of the sum).  We advance this approach to a more general-purpose technique, both by using many subpopulations to refine our estimates and by considering the entire space of possible joint distributions instead of simply a particular aspect of the joint.

@@

Our motivation for this problem arose from looking at Allen Institute cell-type assignment of cells, performed using two different experimental techniques (also called experimental ``modalities'').  The only thing connecting the two modalities was that each modality had a variety of samples drawn from a variety of sub-populations, and the same sub-populations were used for both modalities.  Thus our input was two tables: (technique I cell-type $\times$ sub-population), and (technique II cell-type $\times$ sub-population) (Figure \ref{fig:alleninput}).  A key question immediately appears from these tables: does the way in which we have clustered cells into different types under technique I line up nicely with the way in which we have clustered cells under technique II?  The Markov Link Method gives us a way of addressing this question that is completely agnostic about the underlying genetic markers that were used to produce the clusters (this in constrast to methods such as \cite{kiselev2018scmap} or \cite{crow2018characterizing} which assume both that the clusters arise from observables common to both techniques and that those observables are already at least marginally calibrated between the methods).   Our output was a set $\hat \Theta$ of possible ways that technique I cell-types could line up against the technique II cell-types.  That is, each element of $\hat \Theta$ was itself a table of the form (technique I cell-type $\times$ technique II cell-type).
